cardinalR: Generating interesting high-dimensional data structures

### 2025-09-19 paper-cardinalR.Rmd
Updated Swissroll noise
Updated pyramid section

For all pyramid shapes, if $p > 3$, the remaining $p - 3$ dimensions (i.e., $X_4$ to $X_{p-1}$) are additional noise.

```{r}
pyramid_tb <- tibble(fun = c("gen_pyrrect",
                             "gen_pyrtri",
                             "gen_pyrstar",
                             "gen_pyrholes"), 
                        exp = c("Rectangular-base pyramid, with a sharp or blunted apex.",
          "Triangular-base pyramid, with a sharp or blunted apex.",
          "Star-shaped base pyramid, with a sharp or blunted apex.",
          "Pyramid containing triangular pyramid-shaped holes."))
```

@@

#### `gen_pyrfrac()`

The `gen_pyrfrac(n, p)` function generates $n$ points embedded in a $p\text{-}D$ simplex using a chaotic attractor-like midpoint algorithm (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:pyr) d"} else { "Figure \\@ref(fig:pyr-proj) d"})`). The result is a fractal-like structure that reveals holes or gaps in the data cloud, forming a "Sierpinski-like pyramid" in high dimensions.

Let $X_1, X_2, \dots, X_p$ denote the coordinates of the generated points. The generation process begins with an initial point $T_0 \in [0, 1]^p$ drawn from a uniform distribution: $T_0 \sim U(0, 1)^p$.

Let $C_1, C_2, \dots, C_{p+1}$ denote the corner vertices of a $p\text{-}D$ simplex. At each iteration $i = 1, \dots, n$, a new point is computed by taking the midpoint between the previous point $T_{i-1}$ and a randomly selected vertex $C_k$: $T_i = 1/2(T_{i-1} + C_k), \quad C_k \in \{C_1, \dots, C_{p+1}\}$. This recursive midpoint rule generates self-similar patterns with systematic voids (holes) between clusters of points. The points remain bounded inside the convex hull of the simplex. The final output is a $n \times p$ matrix where each row represents a point: $X = \{T_1, T_2, \dots, T_n\}, \quad X \in \mathbb{R}^{n \times p}$.

@@

To further generalize the Swiss roll structure and introduce realistic noise, we define a function `gen_swissroll(n, p, w)`, where $n$ is the number of points, $p$ is the total number of dimensions, and $w$ is the vertical range in the third dimension (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:swissroll)"} else { "Figure \\@ref(fig:swissroll-proj)"})`). The first three dimensions form the classic $3\text{-}D$ Swiss roll shape. The $X_1 = t \cos(t)$, $X_2 = t \sin(t)$, $X_3 \sim U(w_1, w_2), \text{where } t \sim U(0, 3\pi)$. For $p > 3$, the remaining $p - 3$ dimensions are filled with small Gaussian noise to simulate high-dimensional complexity.

>>>>

For all the above pyramid shapes, if $p > 3$, the remaining $p - 3$ dimensions (i.e., $X_4$ to $X_{p-1}$) are additional noise.

Finally, for the Sierpinski-like pyramid (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:pyr) d"} else { "Figure \\@ref(fig:pyr-proj) d"})`), let $X_1, X_2, \dots, X_p$ denote the coordinates of the generated points. The generation process begins with an initial point $T_0 \in [0, 1]^p$ drawn from a uniform distribution: $T_0 \sim U(0, 1)^p$. Let $C_1, C_2, \dots, C_{p+1}$ denote the corner vertices of a $p\text{-}D$ simplex. At each iteration $i = 1, \dots, n$, a new point is computed by taking the midpoint between the previous point $T_{i-1}$ and a randomly selected vertex $C_k$: $T_i = 1/2(T_{i-1} + C_k), \quad C_k \in \{C_1, \dots, C_{p+1}\}$. This recursive midpoint rule generates self-similar patterns with systematic voids (holes) between clusters of points. The points remain bounded inside the convex hull of the simplex. The final output is a $n \times p$ matrix where each row represents a point: $X = \{T_1, T_2, \dots, T_n\}, \quad X \in \mathbb{R}^{n \times p}$.

```{r}
pyramid_tb <- tibble(fun = c("gen_pyrrect",
                             "gen_pyrtri",
                             "gen_pyrstar",
                             "gen_pyrfrac"), 
                        exp = c("Rectangular-base pyramid, with a sharp or blunted apex.",
          "Triangular-base pyramid, with a sharp or blunted apex.",
          "Star-shaped base pyramid, with a sharp or blunted apex.",
          "Pyramid containing triangular pyramid-shaped holes."))
```

@@

@@

To further generalize the Swiss roll structure and introduce realistic noise, we define a function `gen_swissroll(n, p, w)`, where $n$ is the number of points, $p$ is the total number of dimensions, and $w$ is the vertical range in the third dimension (`r knitr::asis_output(if (knitr::is_html_output()) { "Figure \\@ref(fig:swissroll)"} else { "Figure \\@ref(fig:swissroll-proj)"})`). The first three dimensions form the classic $3\text{-}D$ Swiss roll shape. The $X_1 = t \cos(t)$, $X_2 = t \sin(t)$, $X_3 \sim U(w_1, w_2), \text{where } t \sim U(0, 3\pi)$. For $p > 3$, the remaining $p - 3$ dimensions are filled with noise to simulate high-dimensional complexity.

### 2025-09-19 paper-cardinalR.tex
Updated Swissroll noise
Updated pyramid section

For all pyramid shapes, if \(p > 3\), the remaining \(p - 3\) dimensions (i.e., \(X_4\) to \(X_{p-1}\)) are additional noise.

@@

\caption{\label{tab:pyramid-tb-pdf}cardinalR pyramid data generation functions}
\centering
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{4cm}>{\raggedright\arraybackslash}p{8cm}}
\toprule
Function & Explanation\\
\midrule
gen\_pyrrect & Rectangular-base pyramid, with a sharp or blunted apex.\\
gen\_pyrtri & Triangular-base pyramid, with a sharp or blunted apex.\\
gen\_pyrstar & Star-shaped base pyramid, with a sharp or blunted apex.\\
gen\_pyrholes & Pyramid containing triangular pyramid-shaped holes.\\
\bottomrule
\end{tabular}
\end{table}

@@

\paragraph{\texorpdfstring{\texttt{gen\_pyrfrac()}}{gen\_pyrfrac()}}\label{gen_pyrfrac}

The \texttt{gen\_pyrfrac(n,\ p)} function generates \(n\) points embedded in a \(p\text{-}D\) simplex using a chaotic attractor-like midpoint algorithm (Figure \ref{fig:pyr-proj} d). The result is a fractal-like structure that reveals holes or gaps in the data cloud, forming a ``Sierpinski-like pyramid'' in high dimensions.

Let \(X_1, X_2, \dots, X_p\) denote the coordinates of the generated points. The generation process begins with an initial point \(T_0 \in [0, 1]^p\) drawn from a uniform distribution: \(T_0 \sim U(0, 1)^p\).

Let \(C_1, C_2, \dots, C_{p+1}\) denote the corner vertices of a \(p\text{-}D\) simplex. At each iteration \(i = 1, \dots, n\), a new point is computed by taking the midpoint between the previous point \(T_{i-1}\) and a randomly selected vertex \(C_k\): \(T_i = 1/2(T_{i-1} + C_k), \quad C_k \in \{C_1, \dots, C_{p+1}\}\). This recursive midpoint rule generates self-similar patterns with systematic voids (holes) between clusters of points. The points remain bounded inside the convex hull of the simplex. The final output is a \(n \times p\) matrix where each row represents a point: \(X = \{T_1, T_2, \dots, T_n\}, \quad X \in \mathbb{R}^{n \times p}\).

@@

To further generalize the Swiss roll structure and introduce realistic noise, we define a function \texttt{gen\_swissroll(n,\ p,\ w)}, where \(n\) is the number of points, \(p\) is the total number of dimensions, and \(w\) is the vertical range in the third dimension (Figure \ref{fig:swissroll-proj}). The first three dimensions form the classic \(3\text{-}D\) Swiss roll shape. The \(X_1 = t \cos(t)\), \(X_2 = t \sin(t)\), \(X_3 \sim U(w_1, w_2), \text{where } t \sim U(0, 3\pi)\). For \(p > 3\), the remaining \(p - 3\) dimensions are filled with small Gaussian noise to simulate high-dimensional complexity.

>>>>

For all the above pyramid shapes, if \(p > 3\), the remaining \(p - 3\) dimensions (i.e., \(X_4\) to \(X_{p-1}\)) are additional noise.

Finally, for the Sierpinski-like pyramid (Figure \ref{fig:pyr-proj} d), let \(X_1, X_2, \dots, X_p\) denote the coordinates of the generated points. The generation process begins with an initial point \(T_0 \in [0, 1]^p\) drawn from a uniform distribution: \(T_0 \sim U(0, 1)^p\). Let \(C_1, C_2, \dots, C_{p+1}\) denote the corner vertices of a \(p\text{-}D\) simplex. At each iteration \(i = 1, \dots, n\), a new point is computed by taking the midpoint between the previous point \(T_{i-1}\) and a randomly selected vertex \(C_k\): \(T_i = 1/2(T_{i-1} + C_k), \quad C_k \in \{C_1, \dots, C_{p+1}\}\). This recursive midpoint rule generates self-similar patterns with systematic voids (holes) between clusters of points. The points remain bounded inside the convex hull of the simplex. The final output is a \(n \times p\) matrix where each row represents a point: \(X = \{T_1, T_2, \dots, T_n\}, \quad X \in \mathbb{R}^{n \times p}\).

@@

\caption{\label{tab:pyramid-tb-pdf}cardinalR pyramid data generation functions}
\centering
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{4cm}>{\raggedright\arraybackslash}p{8cm}}
\toprule
Function & Explanation\\
\midrule
gen\_pyrrect & Rectangular-base pyramid, with a sharp or blunted apex.\\
gen\_pyrtri & Triangular-base pyramid, with a sharp or blunted apex.\\
gen\_pyrstar & Star-shaped base pyramid, with a sharp or blunted apex.\\
gen\_pyrfrac & Pyramid containing triangular pyramid-shaped holes.\\
\bottomrule
\end{tabular}
\end{table}

@@

@@

To further generalize the Swiss roll structure and introduce realistic noise, we define a function \texttt{gen\_swissroll(n,\ p,\ w)}, where \(n\) is the number of points, \(p\) is the total number of dimensions, and \(w\) is the vertical range in the third dimension (Figure \ref{fig:swissroll-proj}). The first three dimensions form the classic \(3\text{-}D\) Swiss roll shape. The \(X_1 = t \cos(t)\), \(X_2 = t \sin(t)\), \(X_3 \sim U(w_1, w_2), \text{where } t \sim U(0, 3\pi)\). For \(p > 3\), the remaining \(p - 3\) dimensions are filled with noise to simulate high-dimensional complexity.
